### 2023-06-02 paper/localpsf.tex
passable complete draft

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$.}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$.}
	\end{subfigure}
    \end{center}
	\caption{
		(Top) Bird's eye view of the ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(Bottom left) True parameter, $\basalfriction_\text{true}$. (Bottom right) True velocity $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

Table~\ref{tab:newton_convergence_table} shows the performance of our preconditioner for accelerating the solution of the optimization problem to reconstruct $\mathbf{q}$ from observations with $5\%$ noise. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and reuse it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). Using PSF (5) reduces the total number of Stokes PDE solves substantially compared to regularization preconditioning and no preconditioning\footnote{Interestingly, we observe that no preconditioning outperforms regularization preconditioning.}. In Figure~\ref{fig:stokes_reconstructions} we show reconstructions for $1\%$, $5\%$, and $25\%$ noise.
%
%
\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{stokes_pcg_convergence.pdf}
		\caption{}
		\label{fig:cg_convergence}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
	\begin{center}
		\includegraphics[scale=1.0]{stokes_geigs_figure.pdf}
	\end{center}
	\caption{}
	\label{fig:generalized_eigenvalues}
    \end{subfigure}
	\caption{Left (\ref{fig:cg_convergence}) shows the convergence history for solving $\bm{H} \bm{x} = \bm{b}$ using PCG, where $\bm{b}$ has i.i.d. random entries drawn from the standard normal distribution and $\mathbf{H}$ is evaluated at the solution of the inverse problem. 
%	The reference solution $\bm{x}$ is determined by solving $\bm{H} \bm{x} = \bm{b}$. 
%	Here we use $\gamma=7.3 \times 10^3$, which is determined by the Morozov discrepancy principle when using $5\%$ noise.
	Results in these figures are shown for our preconditioner with 1, 5, and 25 batches (PSF (1), PSF (5), and PSF(25), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using $\bm{H}_\text{gn}$.
	Right (\ref{fig:generalized_eigenvalues}): eigenvalues for generalized eigenvalue problem $\bm{H} \bm{u}_{k}=\lambda_{k} \bm{\preconditioner} \bm{u}_{k}$. Here $\bm{H}$ is the Hessian and $\bm{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\bm{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), or the regularization Hessian (REG).}
	\label{fig:krylov_convergence}
\end{figure} 
%

We build PSF (1), PSF (5), and PSF (25) preconditioners based on the Gauss-Newton Hessian evaluated at the converged solution $\bm{\basalfriction}$. We use PCG to solve a linear system with the Hessian as the coefficient operator and a right hand vector with random i.i.d. entries drawn from the standard normal distribution. In Figure~\ref{fig:cg_convergence} we compare the convergence of PCG for solving this linear system using the PSF (1), PSF (5), PSF (25), REG, and NONE preconditioners. PCG converges fastest with the PSF preconditioners, with PSF (25) converging fastest, followed by PSF (5), followed by PSF (1), as expected. PCG converges much slower with no preconditioning and regularization preconditioning than it does with PSF preconditioning, with no preconditioning outperforming regularization preconditioning. 
In Figure~\ref{fig:generalized_eigenvalues} we show the generalized eigenvalues for the generalized eigenvalue problem
$
\bm{H} \mathbf{u} = \lambda \bm{\preconditioner} \mathbf{u}.
$ 
The matrix $\bm{\preconditioner}$ is one of the PSF (1), PSF (5), or PSF (25) Gauss-Newton Hessian approximations constructed at the converged solution $\bm{\basalfriction}$, or the regularization Hessian. The PSF preconditioners cluster the eigenvalues of the Hessian near one, with more batches yielding better clustering. The regularization preconditioner clusters the trailing eigenvalues but amplifies leading eigenvalues.

@@

Here we consider a time-dependent advection-diffusion equation for
which we seek to infer the unknown spatially varying initial condition, $\ipar$, from observations of the state  
at a final time. This PDE
models diffusive transport in a domain $\advdomain \subset \R^d$, which is
depicted in Figure~\ref{fig:adv_inversion1}. In this case, the state, $\concentration(x,t)$, could be interpreted as the concentration of a contaminant. The problem description
below closely follows the one in~\cite{PetraStadler11}. The domain
boundaries $\partial \advdomain$ include the outer boundaries as well as the
internal boundaries of the rectangles, which represent buildings.  The
parameter-to-observable map $\iFF$ in this case maps an initial
condition $\ipar \in L^2(\advdomain)$ to observations of the concentration at a final time, $\concentration(x, T)$,  through solution of the advection-diffusion
equation given by
\begin{equation}\label{eq:ad}
  \begin{aligned}
    \concentration_t - \kappa\Delta \concentration + v\cdot\nabla \concentration &= 0 & \quad&\text{in
    }\advdomain\times (0,T), \\
    \concentration(\cdot, 0) &= \ipar  &&\text{in } \advdomain , \\
    \kappa\nabla \concentration\cdot \nu &= 0 &&\text{on } \partial \advdomain \times (0,T).
  \end{aligned}
\end{equation}
Here, $\kappa > 0$ is a diffusivity coefficient, $\nu$ is the boundary unit normal vector, and $T > 0$ is the
final time.  The velocity field $v:\advdomain \rightarrow \mathbb{R}^d$, shown in
Figure~\ref{fig:adv_inversion1}, is computed by solving the
steady-state Navier-Stokes equations for a two dimensional flow with
Reynolds number 50 and boundary conditions as in
Figure~\ref{fig:adv_inversion} (left); see~\cite{PetraStadler11} for
details.  The time evolution of the state variable $\concentration$ from a given
initial condition $\ipar$ is illustrated in
Figure~\ref{fig:adv_inversion1}.

@@

\begin{figure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.14]{true_initial_condition.png}
		%		\caption{}
		%		\label{fig:newtoniter2}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.14]{wind_velocity_colorscheme3.png}
		%		\caption{}
		%		\label{fig:newtoniter1}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.14]{noisy_observations_T1.0_K3.2e-4.png}
		%		\caption{}
		%		\label{fig:newtoniter3}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\includegraphics[scale=0.14]{reconstructed_initial_condition_T1.0_K3.2e-4.png}
		%		\caption{}
		%		\label{fig:newtoniter4}
	\end{subfigure}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{(Left) True initial condition. (Middle left) Velocity field. Color indicates magnitude of velocity vector. (Middle right) Noisy observations of concentration at final time. (Right) Reconstructed initial condition.}
	\label{fig:adv_inversion1} 
\end{figure}

\begin{table}
	\centering
	{
		%	\setlength{\tabcolsep}{0.1em}
		%	\renewcommand{\arraystretch}{2.0}
		\begin{tabular}{cccc}
			& $T=0.5$ & $T=1.0$ & $T=2.0$ \\
			$\kappa=1.0e{-4}$  & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K1.e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K1.e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K1.e-4.png} \\
			$\kappa=3.2e{-4}$ & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K3.2e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K3.2e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K3.2e-4.png} \\
			$\kappa=1.0e{-3}$ & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K1.e-3.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K1.e-3.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K1.e-3.png}
		\end{tabular}
	}
	\caption{Impulse responses for a variety of different diffusion parameters $\kappa$ and simulation times $T$.}
	\label{tbl:adv_vary_kappa_tf}
\end{table}

@@

\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{adv_pcg_convergence.pdf}
%		\begin{tikzpicture}[baseline, trim axis right]
%		\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
%		ymode=log,
%		ymin=1e-11,
%		ymax=2,
%		xmax=1000,
%		axis x line*=bottom,
%		axis y line*=left,
%		xtick={0, 200, 400, 600,800,1000},
%		xlabel={CG iteration},
%		ylabel={$\|\bm{x}_k - \bm{x}\|/\|\bm{x}\|$},
%		title={CG convergence},
%		legend style={at={(1.2,1)},anchor=north east},
%		%legend pos=north east,
%		legend style={font=\small}
%		]
%		\addlegendentry{Reg}
%		\addplot[dotted, line width=0.5mm] table {advDiff_cg_rand_errs_reg.dat};
%		\addlegendentry{None}
%		\addplot[dash dot, line width=0.5mm] table {advDiff_cg_rand_errs_none.dat};
%		\addlegendentry{PSF (1)}
%		\addplot[blue, line width=0.5mm] table {advDiff_cg_rand_errs_psf1.dat};
%		\addlegendentry{PSF (5)}
%		\addplot[red, line width=0.5mm] table {advDiff_cg_rand_errs_psf5.dat};
%		\addlegendentry{PSF (25)}
%		\addplot[purple, line width=0.5mm] table {advDiff_cg_rand_errs_psf25.dat};
%		\end{axis}
%		\end{tikzpicture}
		\caption{}
		\label{fig:adv_cg_convergence}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
		\centering
		\includegraphics[scale=1.0]{adv_geigs_figure.pdf}
%		\begin{tikzpicture}[baseline, trim axis right]
%\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
%ymode=log,
%xmin=-50,
%xmax=600,
%ymax=1.e6,
%title={Generalized eigenvalues},
%xlabel={$k$, generalized eigenvalue \#},
%ylabel={$\lambda_{k}$},
%xtick={0, 300, 600, 900, 1200},
%%legend pos=north east,
%legend style={font=\small},
%ylabel near ticks, ylabel shift={-6pt},
%axis x line*=bottom,
%axis y line*=right,
%%legend style={at={(0.95,1)},anchor=north east},
%]
%%\addlegendentry{REG}
%\addplot[dotted, line width=0.5mm] table[x=k, y=reg] {advDiff_generalizedEigenvalues.dat};
%\addplot[dash dot, line width=0.5mm] table[x=k, y=none] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (1)}
%\addplot[blue, line width=0.5mm] table[x=k, y=psf1] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (5)}
%\addplot[red, line width=0.5mm] table[x=k, y=psf5] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (25)}
%\addplot[purple, line width=0.5mm] table[x=k, y=psf25] {advDiff_generalizedEigenvalues.dat};
%\end{axis}
%\end{tikzpicture}
		\caption{}
		\label{fig:adv_generalized_eigenvalues}
	\end{subfigure}
	\caption{Left (\ref{fig:adv_cg_convergence}) shows the convergence history for solving $\bm{H} \bm{x} = \bm{b}$ using PCG, where $\bm{b}$ has i.i.d. random entries drawn from the standard normal distribution. BLAH BLAH BLAH
		%		The y-axis is $\|b-Hx_k\|/\|b\|$, where $x_k$ is the $k$th PCG iterate. 
		Here we use $\gamma=7.88 \times 10^{-6}$, which is determined by the Morozov discrepancy principle. Results in these figures are shown for our preconditioner with 5, 15, and 30 batches (PSF (5), PSF (15), and PSF(30), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using $\bm{H}_\text{gn}$.}
	\label{fig:adv_krylov_convergence}
\end{figure}

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method only requires access to the operator via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes and improves upon classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form an H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a new procedure that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessians in a Stokes ice sheet and advection-diffusion inverse problems, and saw that the approximation substantially outperforms existing Hessian approximation methods. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, our method offers a \emph{data scalable} alternative to conventional low rank approximation, because our method can form high rank approximations of an operator using a small number of operator applies, and thus forward and adjoint PDE solves.

>>>>

\begin{figure}
	\begin{center} 
    \begin{subfigure}{0.99\textwidth}
    \begin{center} 
	\includegraphics[scale=0.23]{meshHeight_view2_edges.png}
	\end{center}
	\caption{Ice sheet model geometry}
	\label{fig:ice_mountain_mesh}
    \end{subfigure} \\
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{mtrue_withColorBar.png}
		\caption{$\basalfriction_\text{true}$.}
		\label{fig:true_beta}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\includegraphics[scale=0.22]{trueVelocity2_glyphs.png}
		\caption{$\velocity_\text{true}$.}
		\label{fig:stokes_velocity}
	\end{subfigure}
    \end{center}
	\caption{
		(\ref{fig:ice_mountain_mesh}) Bird's eye view of the ice sheet, discretized by a mesh of tetrahedra. Color indicates the height of the base of the ice sheet (i.e., the mountain topography). The radius of the domain is $10^4$ meters, the maximum height of the mountain is $2.1 \times 10^3$ meters, and the average thickness of the ice sheet is 250 meters.		
		(\ref{fig:true_beta}) True parameter, $\basalfriction_\text{true}$. (\ref{fig:stokes_velocity}) True velocity $\velocity_\text{true}$. Arrows indicate the direction of $\velocity_\text{true}$ and color indicates the magnitude of $\velocity_\text{true}$.}
	\label{fig:true_beta_u}
\end{figure}

@@

Table~\ref{tab:newton_convergence_table} shows the performance of our preconditioner for accelerating the solution of the optimization problem to reconstruct $\mathbf{q}$ from observations with $5\%$ noise. We build the PSF (5) preconditioner in the third Gauss-Newton iteration, and reuse it for all subsequent Gauss-Newton and Newton iterations. No preconditioning is used in the Gauss-Newton iterations before the PSF (5) preconditioner is built. We compare our method with the most commonly used existing preconditioners: no preconditioning (NONE), and preconditioning by the regularization term in the Hessian (REG). Using PSF (5) reduces the total number of Stokes PDE solves substantially compared to regularization preconditioning and no preconditioning\footnote{Interestingly, we observe that no preconditioning outperforms regularization preconditioning.}. In Figure~\ref{fig:stokes_reconstructions} we show reconstructions for $1\%$, $5\%$, and $25\%$ noise.
%
%
\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{stokes_pcg_convergence.pdf}
		\caption{}
		\label{fig:cg_convergence}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
	\begin{center}
		\includegraphics[scale=1.0]{stokes_geigs_figure.pdf}
	\end{center}
	\caption{}
	\label{fig:generalized_eigenvalues}
    \end{subfigure}
	\caption{Left (\ref{fig:cg_convergence}) shows the convergence history for solving $\bm{H} \bm{x} = \bm{b}$ using PCG, where $\bm{b}$ has i.i.d. random entries drawn from the standard normal distribution and $\mathbf{H}$ is evaluated at the solution of the inverse problem. 
%	The reference solution $\bm{x}$ is determined by solving $\bm{H} \bm{x} = \bm{b}$. 
%	Here we use $\gamma=7.3 \times 10^3$, which is determined by the Morozov discrepancy principle when using $5\%$ noise.
	Results in these figures are shown for our preconditioner with 1, 5, and 25 batches (PSF (1), PSF (5), and PSF(25), respectively), regularization preconditioning (REG), and no preconditioning (NONE). The preconditioner is constructed using $\bm{H}_\text{gn}$.
	Right (\ref{fig:generalized_eigenvalues}): eigenvalues for generalized eigenvalue problem $\bm{H} \bm{u}_{k}=\lambda_{k} \bm{\preconditioner} \bm{u}_{k}$. Here $\bm{H}$ is the Hessian and $\bm{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\bm{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:krylov_convergence}
\end{figure} 
%

We build PSF (1), PSF (5), and PSF (25) preconditioners based on the Gauss-Newton Hessian evaluated at the converged solution $\bm{\basalfriction}$. We use PCG to solve a linear system with the Hessian as the coefficient operator and a right hand vector with random independent and identically distributed (i.i.d.) entries drawn from the standard normal distribution. In Figure~\ref{fig:cg_convergence} we compare the convergence of PCG for solving this linear system using the PSF (1), PSF (5), PSF (25), REG, and NONE preconditioners. PCG converges fastest with the PSF preconditioners, with PSF (25) converging fastest, followed by PSF (5), followed by PSF (1), as expected. PCG converges much slower with no preconditioning and regularization preconditioning than it does with PSF preconditioning, with no preconditioning outperforming regularization preconditioning. 
In Figure~\ref{fig:generalized_eigenvalues} we show the generalized eigenvalues for the generalized eigenvalue problem
$
\bm{H} \mathbf{u} = \lambda \bm{\preconditioner} \mathbf{u}.
$ 
The matrix $\bm{\preconditioner}$ is one of the PSF (1), PSF (5), or PSF (25) Gauss-Newton Hessian approximations constructed at the converged solution $\bm{\basalfriction}$, or the regularization Hessian. The PSF preconditioners cluster the eigenvalues of the Hessian near one, with more batches yielding better clustering. The regularization preconditioner clusters the trailing eigenvalues but amplifies leading eigenvalues.

@@

Here we consider a time-dependent advection-diffusion equation for
which we seek to infer the unknown spatially varying initial condition, $\ipar$, from observations of the state  
at a final time. This PDE
models diffusive transport in a domain $\advdomain \subset \R^d$, which is
depicted in Figure~\ref{fig:adv_inversion1}. In this case, the state, $\concentration(x,t)$, could be interpreted as the concentration of a contaminant. The problem description
below closely follows the one in~\cite{PetraStadler11}. The domain
boundaries $\partial \advdomain$ include the outer boundaries as well as the
internal boundaries of the rectangles, which represent buildings.  The
parameter-to-observable map $\iFF$ in this case maps an initial
condition $\ipar \in L^2(\advdomain)$ to observations of the concentration at a final time, $\concentration(x, T)$,  through solution of the advection-diffusion
equation given by
\begin{equation}\label{eq:ad}
  \begin{aligned}
    \concentration_t - \kappa\Delta \concentration + v\cdot\nabla \concentration &= 0 & \quad&\text{in
    }\advdomain\times (0,T), \\
    \concentration(\cdot, 0) &= \ipar  &&\text{in } \advdomain , \\
    \kappa\nabla \concentration\cdot \nu &= 0 &&\text{on } \partial \advdomain \times (0,T).
  \end{aligned}
\end{equation}
Here, $\kappa > 0$ is a diffusivity coefficient, $\nu$ is the boundary unit normal vector, and $T > 0$ is the
final time.  The velocity field, $v:\advdomain \rightarrow \mathbb{R}^d$, is computed by solving the
steady-state Navier-Stokes equations for a two dimensional flow with
Reynolds number 50, with boundary conditions $v(x)=(0,1)$ on the left boundary, $v(x)=(0,-1)$ on the right boundary, and $v(x)=(0,0)$ on the top and bottom boundaries, as in~\cite[Section 3]{PetraStadler11}. In the numerical results we use a checkerboard image for the initial condition and add $5\%$ multiplicative noise to the observations at the final time, $T$. We use $\kappa=3.2e{-1}$ and $T=1.0$ for all results, except for Table \ref{tab:adv_convergence} and Figure \ref{tbl:adv_vary_kappa_tf} where we vary $\kappa$ and $T$ to study the impact of these parameters on the impulse responses and our preconditioner's effectiveness. The initial condition, velocity field, noisy observations, and reconstructed initial condition are shown in Figure~\ref{fig:adv_inversion1}.

In Table \ref{tab:adv_convergence} we show the number of PCG iterations required to solve the Newton linear system to a relative error tolerance of $10^{-6}$. We show results for $T$ ranging from $0.5$ to $2.0$ and $\kappa$ ranging from $10^{-4}$ to $10^{-3}$, using our PSF preconditioner with $1$, $5$, and $25$ batches, regularization preconditioning, and no preconditioning. PSF preconditioning performs better than regularization preconditioning and no preconditioning, with more batches yielding better results. The PSF preconditioners perform best in the high-rank regime where $T$ and $\kappa$ are small, and the observations are highly informative about the initial condition. This makes sense given the effect of $T$ and $\kappa$ on the shape of the impulse responses, as shown in Figure \ref{tbl:adv_vary_kappa_tf}. As $\kappa$ increases, the impulse responses grow wider and as $T$ increases the impulse responses grow longer. This means that the impulse responses become more local as $\kappa$ and $T$ decrease, so more impulse responses can fit within each batch.

In Figure \ref{fig:adv_cg_convergence} we show the convergence of PCG for solving $\mathbf{H}\mathbf{x}=\mathbf{b}$, where $\mathbf{b}$ has i.i.d. random entries drawn from a standard normal distribution. The preconditioners used, $\widetilde{H}$, are our PSF preconditioner with $1$, $5$, or $25$ batches, regularization preconditioning, and no preconditioning. The PSF preconditioners converge fastest, with more batches yielding faster convergence. In Figure \ref{fig:adv_generalized_eigenvalues}, we show the eigenvalues for the generalized eigenvalue problem $\mathbf{H}\mathbf{u}=\lambda \widetilde{\mathbf{H}}\mathbf{u}$, where the $\widetilde{H}$ are the preconditioners stated above. The PSF preconditioners cluster the eigenvalues of the Hessian near one, and the regularization preconditioner clusters the trailing eigenvalues but amplifies leading eigenvalues. For the PSF preconditioner, more batches yields better clustering.

@@

\begin{figure}
	\begin{center}
	\begin{subfigure}{0.24\textwidth}
		\begin{center}
		\includegraphics[scale=0.14]{true_initial_condition.png}
	\end{center}
				\caption{True $q$}
				\label{fig:adv_initial_condition}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\begin{center}
		\includegraphics[scale=0.14]{wind_velocity_colorscheme3.png}
		\end{center}
				\caption{$v$}
				\label{fig:adv_velocity}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\begin{center}
		\includegraphics[scale=0.14]{noisy_observations_T1.0_K3.2e-4.png}
		\end{center}
				\caption{Noisy observations}
				\label{fig:adv_observations}
	\end{subfigure}
	\begin{subfigure}{0.24\textwidth}
		\begin{center}
		\includegraphics[scale=0.14]{reconstructed_initial_condition_T1.0_K3.2e-4.png}
		\end{center}
				\caption{Reconstructed $q$}
				\label{fig:adv_reconstruction}
	\end{subfigure}
	\end{center}
	%\begin{subfigure}{0.05\textwidth}
	%	\includegraphics[scale=0.1]{mColorBar.png}
	%\end{subfigure}
	\caption{(\ref{fig:adv_initial_condition}) True initial condition. (\ref{fig:adv_velocity}) Velocity field. Color indicates magnitude of velocity vector. (\ref{fig:adv_observations}) Noisy observations of concentration at final time. (\ref{fig:adv_reconstruction}) Reconstructed initial condition.}
	\label{fig:adv_inversion1} 
\end{figure}

@@

\begin{figure}
	\centering
	{
		%	\setlength{\tabcolsep}{0.1em}
		%	\renewcommand{\arraystretch}{2.0}
		\begin{tabular}{cccc}
			& $T=0.5$ & $T=1.0$ & $T=2.0$ \\
			$\kappa=1.0e{-4}$  & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K1.e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K1.e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K1.e-4.png} \\
			$\kappa=3.2e{-4}$ & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K3.2e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K3.2e-4.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K3.2e-4.png} \\
			$\kappa=1.0e{-3}$ & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T0.5_K1.e-3.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T1.0_K1.e-3.png} & \includegraphics[scale=0.15, valign=c]{impulse_response_P(0.3,0.6)_T2.0_K1.e-3.png}
		\end{tabular}
	}
	\caption{Impulse responses for a variety of different diffusion parameters $\kappa$ and simulation times $T$.}
	\label{tbl:adv_vary_kappa_tf}
\end{figure}

\begin{figure}
	\begin{subfigure}{0.52\textwidth}
		\centering
		\includegraphics[scale=1.0]{adv_pcg_convergence.pdf}
%		\begin{tikzpicture}[baseline, trim axis right]
%		\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
%		ymode=log,
%		ymin=1e-11,
%		ymax=2,
%		xmax=1000,
%		axis x line*=bottom,
%		axis y line*=left,
%		xtick={0, 200, 400, 600,800,1000},
%		xlabel={CG iteration},
%		ylabel={$\|\bm{x}_k - \bm{x}\|/\|\bm{x}\|$},
%		title={CG convergence},
%		legend style={at={(1.2,1)},anchor=north east},
%		%legend pos=north east,
%		legend style={font=\small}
%		]
%		\addlegendentry{Reg}
%		\addplot[dotted, line width=0.5mm] table {advDiff_cg_rand_errs_reg.dat};
%		\addlegendentry{None}
%		\addplot[dash dot, line width=0.5mm] table {advDiff_cg_rand_errs_none.dat};
%		\addlegendentry{PSF (1)}
%		\addplot[blue, line width=0.5mm] table {advDiff_cg_rand_errs_psf1.dat};
%		\addlegendentry{PSF (5)}
%		\addplot[red, line width=0.5mm] table {advDiff_cg_rand_errs_psf5.dat};
%		\addlegendentry{PSF (25)}
%		\addplot[purple, line width=0.5mm] table {advDiff_cg_rand_errs_psf25.dat};
%		\end{axis}
%		\end{tikzpicture}
		\caption{}
		\label{fig:adv_cg_convergence}
	\end{subfigure}
	\begin{subfigure}{0.47\textwidth}
		\centering
		\includegraphics[scale=1.0]{adv_geigs_figure.pdf}
%		\begin{tikzpicture}[baseline, trim axis right]
%\begin{axis}[width=0.99\linewidth, height=1.0\linewidth,
%ymode=log,
%xmin=-50,
%xmax=600,
%ymax=1.e6,
%title={Generalized eigenvalues},
%xlabel={$k$, generalized eigenvalue \#},
%ylabel={$\lambda_{k}$},
%xtick={0, 300, 600, 900, 1200},
%%legend pos=north east,
%legend style={font=\small},
%ylabel near ticks, ylabel shift={-6pt},
%axis x line*=bottom,
%axis y line*=right,
%%legend style={at={(0.95,1)},anchor=north east},
%]
%%\addlegendentry{REG}
%\addplot[dotted, line width=0.5mm] table[x=k, y=reg] {advDiff_generalizedEigenvalues.dat};
%\addplot[dash dot, line width=0.5mm] table[x=k, y=none] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (1)}
%\addplot[blue, line width=0.5mm] table[x=k, y=psf1] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (5)}
%\addplot[red, line width=0.5mm] table[x=k, y=psf5] {advDiff_generalizedEigenvalues.dat};
%%\addlegendentry{PSF (25)}
%\addplot[purple, line width=0.5mm] table[x=k, y=psf25] {advDiff_generalizedEigenvalues.dat};
%\end{axis}
%\end{tikzpicture}
		\caption{}
		\label{fig:adv_generalized_eigenvalues}
	\end{subfigure}
	\caption{Left (\ref{fig:adv_cg_convergence}) shows the convergence history for solving $\bm{H} \bm{x} = \bm{b}$ using PCG, where $\bm{b}$ has i.i.d. random entries drawn from the standard normal distribution. Right (\ref{fig:adv_generalized_eigenvalues}) shows eigenvalues for generalized eigenvalue problem $\bm{H} \bm{u}_{k}=\lambda_{k} \bm{\preconditioner} \bm{u}_{k}$. Here $\bm{H}$ is the Hessian and the preconditioner $\bm{\preconditioner}$ is the PSF approximation for 1, 5, or 25 batches (PSF (1), PSF (5), and PSF (25), respectively), the regularization Hessian (REG), or the identity matrix (NONE).}
	\label{fig:adv_krylov_convergence}
\end{figure}

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method only requires access to the operator via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form an H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a new procedure that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessians in Stokes ice sheet and advection-diffusion inverse problems, and saw that the approximation substantially outperforms existing Hessian approximation methods. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, our method offers a \emph{data scalable} alternative to conventional low rank approximation, because our method can form high rank approximations of an operator using a small number of operator applies, and thus forward and adjoint PDE solves.
